---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.


## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

```{r}
pg_overdose <- read.csv("data/prince_georges_2022_overdoses.csv")

library(tidyverse)
library(lubridate)
library(ggthemes)
library(tidycensus)

pg_overdose <- pg_overdose |> 
 mutate(date = date(datetime),
    week = week(datetime),
    month = month(datetime, label=TRUE))
```


## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.** The pattern of the results from the pre lab were more clear than the ones in the lab, because the months with the most overdoses in the pre lab were the warmer months, but the months with the most overdoses in PG county are December and July. We connected them by the fact that there tend to be more holidays in the months with the most overdoses, but it's difficult to be certain.

```{r}
calls_per_month <- pg_overdose |> 
  group_by(month) |> 
  summarise(count = n()) |> 
  mutate(percentage = count/sum(count)*100) |> 
  arrange(desc(percentage))

calls_per_month

calls_per_month |> 
  ggplot() +
  geom_bar(aes(x=reorder(month, percentage), weight=percentage)) +
  coord_flip() +
  theme_minimal() +
  labs(
    title="Overdoses Spike in Festive Months",
    x = "Month",
    y = "Percentage of Total Calls",
    caption = "source: Prince George's County EMS")
```

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** The pattern is pretty consistent throughout except in two places where there is a spike, one being in the middle of the year and the other being at the end of the year. There is a huge drop at the end of the line that we identified as an outlier. We found that there was a 53rd week that only had four overdose calls. We think it was most likely a shorter week, because it cam at the very end of the year.


```{r}
calls_per_week <- pg_overdose |> 
  group_by(week) |> 
  summarise(count=n())

calls_per_week

calls_per_week |> 
  ggplot() +
  geom_line(aes(x=week, y=count)) +
  theme_minimal() +
  labs(
    title="Overdoses Spike in the Middle and End of the Year",
    x = "Week",
    y = "Total Calls",
    caption = "source: Prince George's County EMS")
  
```

**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  What this scatterplot is showing is that there is an inverse relationship between median income and the number of 911 calls where as median income goes up, the number of calls goes down. Because this is a correlation and not a very strong one, I wouldn't report it as fact that places with lower income definitely have more overdoses, especially when there are plenty of zip codes where the median income is low that don't have a ton of overdose calls.

```{r}
calls_per_zip <- pg_overdose |> 
  group_by(zipcode) |> 
  summarise(count=n())
calls_per_zip

vacs5 <- load_variables(2019, "acs5", cache = TRUE)

medic_md <- get_acs(
  geography = "zcta",
  variables = c(medincome = "B19013_001"),
  state = "MD",
  year = 2019)
medic_md

od_medinc <- calls_per_zip |> inner_join(medic_md, by=c('zipcode' = 'GEOID'))

od_medinc |> 
  ggplot() +
  geom_point(
    aes(x= estimate, y=count)) +
  theme_minimal() +
  labs(
    title="More 911 Overdose Calls in Zip Codes with Lower Median Income",
    x = "Median Househole Income",
    y = "Number of Calls",
    caption = "source: Prince George's County EMS | Census Bureau") +
  geom_text(aes(x=estimate, y=count+5, label=zipcode), check_overlap = TRUE)
```
