---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564)
# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

cleaned_payments <- payments |>
  clean_names()

cleaned_payments

cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor)

cleaned_md_grants_loans
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1: I filtered the data by "PRINCE GEORGES COUNTY PUBLIC SCHOOL", "PG COUNTY PUBLIC SCHOOL" and "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD". I found 37 records.

```{r}
pg_schools <- cleaned_payments |> 
  filter(str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD")) 

pg_schools
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: The fiscal year that had the largest total was 2022, with $1.6 billion. The aspect that stands out is that 2009 and 2008 is higher than more recent years, I'm wondering if that has anything to do with the recession that was happening at the time.

```{r}
pg_schools |> 
  group_by(fiscal_year) |> 
  summarize(total = sum(amount)) |> 
  arrange(desc(total))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. WHITING TURNER CONTRACTING CO and WHITING TURNER CONTRACTING COMPANY; A MORTON THOMAS & ASSOC INC and A MORTON THOMAS & ASSOCIATES INC; A&I INC and A&I INC.

To clean this, I would use the "case when" function to group the similar names together.

```{r}
cleaned_payments |>
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) |> 
  group_by(vendor_name) |> 
  summarize(total = sum(amount)) |> 
  arrange(vendor_name)
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: I think stem cell research funding could be related to helping to understand diseases, so maybe the huge amount of funding in 2009 and the increase in 2019 and 2020 could be due to the flu and COVID pandemics respectively.

```{r}
cleaned_md_grants_loans |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description)) |> 
  filter(str_detect(source, "STEM CELL")) |> 
  group_by(fiscal_year) |> 
  summarize(total=sum(amount)) |> 
  arrange(fiscal_year)
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 18 rows, which should be 18 grantees, but there are some that seem repeated, for example "UNIVERSITY OF MARYLAND", "University of Maryland College Park" and "UNIVERSITY OF MD,COLLEGE PARK" all seem like they mean the same thing. In order to clean the results, you would need to look for grantees that have the same name and change them to be written in the same.
The most interesting grantee to me is the UMD Data Center, because I take a lot of classes in data analysis and I didn't know what the Data Center was.

```{r}
cleaned_md_grants_loans |> 
mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |>
  mutate(zip5 = case_when(
    zip5 == "Vario" ~ NA,
    zip5 == "UB7 O" ~ NA,
    zip5 == "UB7 " ~ NA,
    .default = zip5
  )) |> 
  filter(str_detect(zip5, "20742")) |> 
  group_by(grantee) |> 
  summarize(total = sum(amount)) |> 
  arrange(desc(total))
```

