---
title: "lab_07"
author: "derek willis"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2021 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
vacs5 <- load_variables(2021, "acs5", cache = TRUE)

vacs5
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? You will need to find the variable for median age first. 
**A1** The Maryland jurisdiction with the lowest median age is Baltimore City with 35.7.

```{r}
medage_md21 <- get_acs(geography="county",
                    variables = c(medage = "B01002_001"),
                    state="MD",
                    year=2021) |> arrange(estimate)
medage_md21
```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. 
**A2** The counties with a median age of at least 50 are Worchester and Talbot counties.

```{r}
medage_md21 |> 
  filter(estimate >= 50)
```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1. Load the Montgomery County 911 overdose calls data.
2. Create a new dataframe with the total number of calls for each zip code.
3. Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one).
4. Get from the ACS data from all Maryland zip codes using that variable.
5. Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
6. Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.

Which zip code has the highest rate of calls? Which has the lowest?

**A3** The zip code with the highest rate of calls is 20877. The one with the lowest rate of calls is 20707.

```{r}
#Part 1
overdose <- read_csv("data/montgomery_2022_overdoses.csv")
```

```{r}
#Part 2
calls_per_zip <- overdose |> 
  group_by(zip) |> 
  summarize(number_of_calls = n())
calls_per_zip

#Parts 3 and 4
over18_md <- get_acs(
  geography = "zcta",
  state = "MD",
  variables = c(adult = "B09021_001"),
  year = 2019
)
over18_md

#Part 5
calls_per_zip$zip <- as.character(calls_per_zip$zip)
calls_per_zip

over18_moco <- over18_md |> left_join(calls_per_zip, join_by(GEOID == zip)) |> drop_na()
over18_moco <- over18_moco |> rename(number_of_adults = estimate)
over18_moco

#Part 6
over18_moco <- over18_moco |> 
  mutate(rate_per_1k = (number_of_calls/number_of_adults)*1000) |> 
  arrange(desc(rate_per_1k))
over18_moco
```

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4** The zip code with the highest rate of oversdose calls is located in Gaithersburg, and the one with the lowest is located in Laurel. The Gaithersburg zip has a median household income of 77K, a 15.4% poverty rate and a 77.8% high school graduation rate. The Laurel zip has a median household income of 91K, a 8.4% poverty rate and a 90.3% high school graduation rate. There could be a story here about the link that poverty and education has with overdoses.

**Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?

**A5** I did get the same two zip codes as before. This gives me more confidence in my previous answers, because the margin of error not being big enough to change the results shows that the results are significant.

```{r}
over18_moco <- over18_moco |>
  mutate(pop_min = number_of_adults - moe) |> 
  mutate(pop_max = number_of_adults + moe) |> 
  mutate(min_1k = (number_of_calls/pop_min)*1000) |> 
  mutate(max_1k = (number_of_calls/pop_max)*1000) |> 
  arrange(desc(min_1k))
over18_moco

over18_moco |> 
  arrange(desc(max_1k))
over18_moco
```

