---
title: "lab_10"
author: "Derek Willis"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries
library(tidyverse)
library(sf)
library(janitor)
library(tidycensus)
library(tigris)
```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history last November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.** Queen Anne's, Carroll and Kent counties all have the biggest diferences between Moore and Brown, with Moore winning more than 7 percentage points than Brown. The question would be about why some people who voted for Moore in these counties didn't also vote for Brown.

```{r}
md_statewide_county <- read_csv("data/md_statewide_county.csv") |> clean_names()

# Election Data
md_statewide_county <- md_statewide_county |> 
  mutate(gov_total = cox + moore + lashar + wallace + harding + sefcik + gov_writein) |> 
  mutate(ag_total = peroutka + brown + ag_writein) |> 
  mutate(moore_pct = (moore/gov_total)*100) |> 
  mutate(brown_pct = (brown/ag_total)*100) |> 
  mutate(difference = moore_pct - brown_pct) |> 
  mutate(geoid = as.character(geoid))

# Geometry
counties <- counties()

md_counties <- counties |>
  filter(STATEFP == "24")

md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD",
              geometry = TRUE)

md_county_population <- md_county_population |>
  rename(county = NAME) |>
  mutate(county = str_to_title(str_remove_all(county,", Maryland"))) |>
  mutate(county = str_trim(county,side="both")) |>
  rename(population = estimate) |>
  select(GEOID, county, population, geometry)

# Joining
election_geometry <- md_county_population |> left_join(md_statewide_county, join_by(GEOID == geoid))

# Map
county_centroids <- st_centroid(md_counties)
county_centroids_df <- as.data.frame(st_coordinates(county_centroids))
county_centroids_df$NAME <- county_centroids$NAME

ggplot() +
  geom_sf(data=election_geometry, aes(fill=difference)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE) +
  theme_minimal()
```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.** The most alarming result is definitely that the rate for power outages in Talbot County significantly higher than the rest. If we had access to data from other months, I'd try to see if this is a long-term trend or not. Then we'd have to talk to people in the county who have been affected by power outages and get their thoughts. 

```{r}
outages <- read_csv("data/county_outages_march2023.csv")

# Rates
outages <- outages |> 
  mutate(rate = (outage_customers/total_customers)*10000)

# Joining
md_county_population <- md_county_population |>
  mutate(county = str_to_title(str_remove_all(county,"County"))) |>
  mutate(county = str_trim(county,side="both"))

outages_counties <- md_county_population |> left_join(outages, join_by(county))

# Map
county_centroids <- st_centroid(md_counties)
county_centroids_df <- as.data.frame(st_coordinates(county_centroids))
county_centroids_df$NAME <- county_centroids$NAME

ggplot() +
  geom_sf(data=outages_counties, aes(fill=rate)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE) +
  scale_fill_viridis_b(option="magma",trans = "log") +
  theme_minimal()
```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.** For Map 1, the two counties that stick out are Caroline and Dorchester, because Caroline had the highest percent change with 426% and Dorchester had the lowest with 104%. An interesting story would be to compare these counties and understand what factors are causing the change. For example, is income a defining factor here?

For Map 2, the counties that stand out are Montgomery and Howard for having the highest rates of electric and hybrid vehicles per 10,000, as well as Somerset for having the smallest rate. The percent change map showed the increase in these types of vehicles while this one shows the rate of ownership. Montgomery and Howard counties are known to have high median incomes, but they ranked lower in the percent change. So maybe there was already an abundance of electric and hybrid vehicles prior to July 2020, so much so that there would have be a massive increase in the last three years for there to be any significant percentage change.

```{r}
electric <- read_csv("data/ev_2020_2023.csv")

# Percent Change
electric <- electric |> 
  mutate(pct_change = ((march_2023-july_2020)/july_2020)*100)

# Joining
electric_counties <- md_county_population |> left_join(electric, join_by(county))

# Map 1
county_centroids <- st_centroid(md_counties)
county_centroids_df <- as.data.frame(st_coordinates(county_centroids))
county_centroids_df$NAME <- county_centroids$NAME

ggplot() +
  geom_sf(data=electric_counties, aes(fill=pct_change)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE) +
  scale_fill_viridis_b(option="magma",trans = "log") +
  theme_minimal()

# Rate
# I've already been using the Maryland population data from the ACS from the beginning, because I found that it had more accurate geometry for the counties than the counties() function in tigris. So I don't need to join it again since I already joined it above in "electric_counties."

electric_counties <- electric_counties |> 
  mutate(rate = (march_2023/population)*10000)

# Map 2
ggplot() +
  geom_sf(data=electric_counties, aes(fill=rate)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE) +
  scale_fill_viridis_b(option="magma",trans = "log") +
  theme_minimal()
```
